On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-17 paper/application.tex
regular update

\section{Application to protoplanetary disks}\label{2dppd}
We now consider the stability of realistic, self-gravitating
protoplanetary disks. Specifically, we will apply our linear framework
to the gravito-turbulent disk model recently developed by 
\cite{rafikov15}. This 2D, Keplerian disk orbits a Solar mass star and 
is defined by the following parameters,

@@

\subsection{PPD cooling}
Energy loss in \citeauthor{rafikov15}'s model is given
by 
\begin{align}\label{real_cool}
  \Lambda = \frac{2\sigma}{f(\tau)}\left(T^4 - T_\mathrm{irr}^4\right)
\end{align}
per unit area, 
\begin{align}
  f(\tau) = \tau + \frac{1}{\tau},
\end{align}
and
\begin{align}
  \tau = \kappa_d(T)\Sigma
\end{align}
is the optical depth and recall Eq. \ref{opacity_law} 
is our opacity model. 
%and $\kappa_d$ is the opacity. 
%We parameterize
%\begin{align}
%  \kappa_d = \kappa_{d0} \Sigma^a T^b,
%\end{align}
%where $\kappa_{d0}, a, b$ are constants.

Since we are interested in the linear response of the system, the
corresponding parameters $\beta$ and $\theta$ are found by
comparing the linearized form of the cooling function
(Eq. \ref{real_cool}) to that of the model function
(Eq. \ref{beta_cool}).

%{\bf as opposed to comparing their unperturbed forms, as done in literature}

Linearizing Eq. \ref{real_cool} gives   
\begin{align}\label{linear_real_cool}
  (\gamma-1)\frac{\delta\Lambda}{\Sigma} = \frac{2\sigma(\gamma-1)
    T^4C_1}{f(\tau)c_{s}^2\Sigma}\left(\frac{\delta P}{\Sigma} -
  \frac{C_2}{C_1}c_{s}^2\frac{\delta\Sigma}{\Sigma}\right), 
\end{align}
where
\begin{align}
C_1(\tau, T) &= 4 - b\times g(\tau, T),\\ 
C_2(\tau, T) &= 4 + (1-b)\times g(\tau, T),\\ 
  g(\tau, T) &= \left( \frac{\tau^2-1}{\tau^2+1}\right)\left(1 -
  \frac{T_\mathrm{irr}^4}{T^4}\right). 
\end{align}
Comparing Eq. \ref{linear_real_cool} with Eq. \ref{beta_cool}, we identify
\begin{align}
  &\beta = \frac{f(\tau)c_s^2\Sigma\Omega}{2\sigma(\gamma-1)C_1T^4},\\
  &\theta = \frac{C_2}{C_1}, 
\end{align}
to be used in the 2D linear framework.

@@

%{\bf caution in applying numerical results obtained from beta cooling to
%systems modeled with real cooling. above example shows that to model
%real, non-irradiated disk with beta cooling actually requires irradiation in
%the latter.}

%Notice that $\theta\to 1$ as 
%$T\toT_\mathrm{irr}$ the disk becomes passive.

\subsection{Inviscid stability condition} 
%We specialize to $a=0, b=2$ for the opacity law. 
We note that for the adopted opacity law, Eq. \ref{opacity_law}, 
\begin{align*}
  \theta = \frac{4-g}{4-2g},
\end{align*}
implying $5/6<\theta<3/2$. 
%For a non-irradiated disk, $T_\mathrm{irr}=0$, $\theta\to5/6$ as
%$\tau\to0$ and $\theta\to3/2$ as $\tau\to\infty$.  
From the results in \S\ref{2d_inviscid}, we conclude that this
system cannot be unstable anywhere without
viscous effects acting on the perturbations if 
\begin{align}
  \gamma > \frac{3}{2} \quad \text{and} \quad Q > \sqrt{\frac{6}{5}} \label{ppd_invisc_cond}
\end{align} 
are both satisfied.

A PPD becomes optically-thin ($\tau\to 0$) at
large radii, and $\theta$ decreases. The first inequality in
Eq. \ref{ppd_invisc_cond} then relaxes to $\gamma > 5/6$ in the
outer disk, which is always true. On the other hand, gravito-turbulent
simulations typically show $Q\simeq 1.5$---2, which satisfies the
second critrion. This means that in the
outer regions of a realistic PPD, cooling itself cannot cause instability.

@@

\subsection{Example 2D calculation}\label{pp2d_example}
We consider a disk with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
$\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. For the adopted opacity
law we follow \citeauthor{rafikov15} and take $\kappa_{d0} =
5\times10^{-4}\mathrm{cm}^2\,\mathrm{g}^{-1}\,\mathrm{K}^{-2}$. Here we use
$\gamma=1.6$ so the inviscid stability condition
(Eq. \ref{ppd_invisc_cond}) is satisfied. Then cooling alone cannot
destabilize the disk. Fig. \ref{rafikov_model} shows the equilirbium
disk profile in terms of $Q$, $\alpha$, $\beta$, and $\theta$.  
These are set as input to the 2D dispersion relation.

>>>>

\section{Application to protoplanetary disks}\label{2dppd}
We now apply our linear framework to assess the stability of realistic 
protoplanetary disks. We consider the gravito-turbulent disk models recently
developed by \cite{rafikov15}. This 2D, Keplerian disk orbits a Solar 
mass star and is defined by the following parameters,

@@

\subsection{PPD cooling}
Energy loss in \citeauthor{rafikov15}'s model is given by 
\begin{align}\label{real_cool}
  \Lambda = \frac{2\sigma}{f(\tau)}\left(T^4 - T_\mathrm{irr}^4\right)
\end{align}
per unit area, 
\begin{align}
  f(\tau) = \tau + \frac{1}{\tau}, \label{ftau} 
\end{align}
and
\begin{align}
  \tau = \kappa_d(T)\Sigma
\end{align}
is the optical depth and recall Eq. \ref{opacity_law} 
is our opacity model. Eq. \ref{ftau} accounts for cooling in the
optically-thin ($\tau\ll 1$) and optically-thick ($\tau\gg1$)
regimes.

Since we are interested in the linear response of the system, the
corresponding cooling parameters $\beta$ and $\theta$ are found by
comparing the linearized form of the cooling function
(Eq. \ref{real_cool}) to that of the model function
(Eq. \ref{beta_cool}).  
%{\bf as opposed to comparing their unperturbed forms, as done in literature}
Linearizing Eq. \ref{real_cool} gives   
\begin{align}\label{linear_real_cool}
  (\gamma-1)\frac{\delta\Lambda}{\Sigma} = \frac{2\sigma(\gamma-1)
    T^4C_1}{f(\tau)c_{s}^2\Sigma}\left(\frac{\delta P}{\Sigma} -
  \frac{C_2}{C_1}c_{s}^2\frac{\delta\Sigma}{\Sigma}\right), 
\end{align}
where
\begin{align}
C_1(\tau, T) &= 4 - b\times g(\tau, T),\\ 
C_2(\tau, T) &= 4 + (1-b)\times g(\tau, T),\\ 
  g(\tau, T) &= \left( \frac{\tau^2-1}{\tau^2+1}\right)\left(1 -
  \frac{T_\mathrm{irr}^4}{T^4}\right). 
\end{align}
Comparing Eq. \ref{linear_real_cool} with Eq. \ref{beta_cool}, we identify
\begin{align}
  &\beta = \frac{f(\tau)c_s^2\Sigma\Omega}{2\sigma(\gamma-1)C_1T^4},\\
  &\theta = \frac{C_2}{C_1}, 
\end{align}
to be used in the 2D linear framework.

@@

\subsection{Inviscid stability condition}
We note that for the adopted opacity law, Eq. \ref{opacity_law}, 
\begin{align*}
  \theta = \frac{4-g}{4-2g},
\end{align*}
implying $5/6<\theta<3/2$. 
%For a non-irradiated disk, $T_\mathrm{irr}=0$, $\theta\to5/6$ as
%$\tau\to0$ and $\theta\to3/2$ as $\tau\to\infty$.  
From the results in \S\ref{2d_inviscid}, we conclude that this
system cannot be unstable anywhere without
viscous effects acting on the perturbations if 
\begin{align}
  \gamma > \frac{3}{2} \quad \text{and} \quad Q >
  \sqrt{\frac{6}{5}} \label{ppd_invisc_cond} 
\end{align} 
are both satisfied.

PPDs become optically-thin ($\tau\to 0$) with 
increasing distance from its host star. Then $\theta$ decreases, and
the first inequality in Eq. \ref{ppd_invisc_cond} relaxes to $\gamma >
5/6$ in the (gravito-turbulent) outer disk, which is always satisfied.  
On the other hand, numerical simulations of gravito-turbulence
show that the second inequality is generally satisfied \citep{rice11}.  
Taken together, this means that the outer regions of a realistic PPD,
cooling itself cannot cause instability.

@@

\subsection{Example 2D calculation}\label{pp2d_example}
We consider a disk with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
$\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. For the adopted opacity
law we follow \citeauthor{rafikov15} and take $\kappa_{d0} =
5\times10^{-4}\mathrm{cm}^2\,\mathrm{g}^{-1}\,\mathrm{K}^{-2}$. Here we use
$\gamma=1.6$ so the inviscid stability condition
(Eq. \ref{ppd_invisc_cond}) is satisfied
everywhere. Fig. \ref{rafikov_model} shows the equilirbium 
disk profile in terms of $Q$, $\alpha$, $\beta$, and $\theta$.  
These are set as input to the 2D dispersion relation.

### 2016-01-17 paper/results2d.tex
regular update

Fig. \ref{gammie_maxrate_plot} show the maximum growth rate as a
function of the cooling time $\beta$ for $\theta=0$. 
%We set $\alpha_b=\theta=0$ and 
%consider three cases (top to bottom panels): 
%\begin{enumerate}[(a)]
%  \item $\alpha$, $\beta$ and $Q$ are independent; \label{casea}
%  \item $\alpha=\alpha(\beta)$ from thermal equlibrium
%    (Eq. \ref{alpha_beta_relation}) and $Q$ specified separately; \label{caseb}
%  \item $\alpha(\beta)$ from Eq. \ref{alpha_beta_relation} and
%    assuming $Q = Q_\mathrm{crit}/\sqrt{\alpha}$. This mimics a
%    basic state maintained by 
%    gravito-turbulence where the viscosity coefficient $\alpha \sim
%    Q^{-2}$ \citep{lin87}.
%    \label{gvisc}
%\end{enumerate} 
There is good agreement between numerical results and
Eq. \ref{gammie_maxrate_simple} for $\beta \gtrsim 1$. 
%The assumption
%of $|\beta S|\ll 1$ is not well satisfied at smaller $\beta$ which
%explains the mismatch there. 
Eq. \ref{gammie_maxrate_simple} gives the
limiting behavior for this case as   
\begin{align*}
  S_*\propto \begin{cases}
%    \frac{\alpha^{1/4}}{Q\beta^{3/4}} \propto \beta^{-3/2} 
    \alpha^{1/4}\beta^{-3/4}Q^{-1}\propto \beta^{-3/2} 
    &  \beta \gg \alpha, \\
%    \frac{1}{Q\sqrt{\alpha}}  
    \alpha^{-1/2}Q^{-1} = \mathrm{const.} & \beta \ll \alpha,    
  \end{cases}
\end{align*}
where we have applied Eq. \ref{alpha_beta_relation}
and \ref{Qalpha} to obtain the final result. 
The disk is formally unstable for all $\beta$, but
the growth timescales are long ($>10$ orbits) for $\beta \gtrsim
20$. This region is marked by the vertical dashed-dotted line in
Fig. \ref{gammie_maxrate_plot}.

%As $\beta\to 0$ the required viscous heating to set up an equilibrium
%state diverges ($\alpha\to\infty$). 
%In the present model the increased
%viscous damping, which is stabilizing, is balanced by the decreasing
%(and hence destablizing) $Q$ required to achieve large $\alpha$
%(Eq. \ref{Qalpha}).

>>>>

Fig. \ref{gammie_maxrate_plot} shows the maximum
growth rate (top panel) and the corresponding wavenumber (bottom
panel) as a function of the cooling time $\beta$ for $\theta=0$.  
There is good agreement between numerical growth rates and
Eq. \ref{gammie_maxrate_simple} for $\beta \gtrsim 1$. 
%The assumption
%of $|\beta S|\ll 1$ is not well satisfied at smaller $\beta$ which
%explains the mismatch there. 
Eq. \ref{gammie_maxrate_simple} gives the
limiting behavior for this case as   
\begin{align*}
  S_*\propto \begin{cases}
%    \frac{\alpha^{1/4}}{Q\beta^{3/4}} \propto \beta^{-3/2} 
    \alpha^{1/4}\beta^{-3/4}Q^{-1}\propto \beta^{-3/2} 
    &  \beta \gg \alpha, \\
%    \frac{1}{Q\sqrt{\alpha}}  
    \alpha^{-1/2}Q^{-1} = \mathrm{const.} & \beta \ll \alpha,    
  \end{cases}
\end{align*}
where we have applied Eq. \ref{alpha_beta_relation}
and \ref{Qalpha}. The disk is formally unstable for all $\beta$, but
growth timescales are long ($>10$ orbits) for $\beta \gtrsim
20$. This region is marked by the vertical dashed-dotted line in
Fig. \ref{gammie_maxrate_plot}. The optimum wavenumber decreases with
the cooling time for $\beta\lesssim O(1)$ because larger scales are
more resistant to the associated increase in viscous damping.

### 2016-01-17 paper/results3d.tex
regular update

\section{Three-dimensional disks with beta cooling}\label{3ddisk}
We briefly confirm the above results in 3D disks with vertical
structure. This is expected to weaken gravitational instabilities
because the disk mass is spread across the vertical dimension. It is
possible to incorportate this effect in the previous 2D framework, but this
introduces additional parameters. We discuss this in Appendix
\ref{3dcorr}. It is more direct to solve the 3D eigenvalue problem to
avoid such uncertainties. 
%We will find that while softening the
%self-gravity in 2D can produce qualitatively similar results to 3D, a
%quantitative match is

>>>>

\section{Three-dimensional disks with beta cooling}\label{3ddisk}
We confirm the above results in 3D disks with vertical
structure. This is expected to weaken gravitational instabilities
because the disk mass is spread across the vertical dimension. It is
possible to incorportate this effect in the previous 2D framework, but this
introduces additional parameters as discussed in Appendix
\ref{3dcorr}. It is more direct to solve the 3D eigenvalue problem to
avoid such uncertainties. 
%We will find that while softening the
%self-gravity in 2D can produce qualitatively similar results to 3D, a
%quantitative match is
